### 2020-08-06 paper/sex_selection_spacing-ver4.tex
Edited fertility section

% For word counting uncomment line below and raggedright below
% http://countwordsfree.com/#
% \usepackage[nolists]{endfloat}

@@

% For word counting remove comment on the line below
% \raggedright

@@

@@

@@

The tempo effect from longer birth intervals means that the total fertility rate may 
underestimate cohort fertility. 
The next question I address is, therefore, to what extent did the changes bias the 
fertility estimates for India? 
To this end, I compare fertility based on a variation of the total fertility rate with 
predicted cohort fertility from the hazard model results.

@@

Table \ref{tab:fertility} shows the two fertility measures by urban/rural and education.
Because the hazard model predictions only use births up to parity four, I calculate
fertility rates using only the same births.
Hence, my fertility rates are not directly comparable to those in the NFHS reports,
but otherwise, I follow the procedure used in the NFHS reports \citep{Croft2018}.
I calculate age-specific fertility rates for five-year age groups using the number of 
births in months 1 to 36 before the survey month and the number of women in each age 
group.

@@

Because NFHS-1 was after the introduction of sex selection, it is not possible 
to calculate a fertility rate in precisely the same manner for a period 
before sex selection was widely available.
Instead, I calculate the fertility rates for women who were between 15 and 39 years of
age five years before the survey month, again using the number of births three years 
before that, shown as ``1987--88''.
Given the relatively low number of births to women 40 to 45 years of age, this approach 
provides the best estimate of the fertility rate at the time when sex selection still was 
not widespread.

To predict cohort fertility based on the hazard models above, I estimate the probability 
of progression to the next parity for each spell.
Since the parity progression depends on the sex composition for prior children for parity 
two and above, I estimate the progression probability for each possible sex composition
and weigh those probabilities with the likelihood that they will occur based on the
prior spells.

Sterilizations are not incorporated into the hazard model because most occur
immediately after giving birth. 
To compensate, I estimate the likelihood of sterilization using a Logit model, predict 
the probability of not getting sterilized, and then use that to scale down the parity
progression probability when predicting cohort fertility.

I include the spell from marriage to first birth, despite the problems
capturing the exact timing of marriages since the estimated progression
probabilities should not be affected by this problem.
To ensure that the predictions depend solely on sample composition and the estimate, I
use the age at marriage for each woman and predict the likelihood of progressing to
each parity, assuming three years increases in age between each parity.
Shorter assumed increases in age between births leads to slightly higher predicted 
fertility.

The survey rounds do not coincide directly with the periods used for the hazard model.
I, therefore, use the model results for 1972--1984, 1985--1994, 1995--2004,
and 2005--2016 for rounds 1 through 4 of the NFHS, respectively.

% Because of the small number of women who had the highest level of education in the
% 1972--1984 period, I do not predict fertility for this case.

The predicted cohort fertility based on the hazard model is higher than the four-parity 
fertility rate in almost all cases.
Only for women with no education in the first period is there almost no difference 
between the two fertility measures; a situation where fertility is high, spacing
very short and likely unchanged for an extended period before.

Consistent with a more substantial bias in the fertility rate when the age of marriage 
and the length of birth intervals are increasing, the absolute bias is least in the first and the last 
period and highest in the middle two periods.
Hence, the fertility rate declined too fast from the mid-1980s to the century's end.
Only recently, as the rate of increase for the birth intervals have
slowed, are the predicted cohort fertility and the fertility beginning
to converge again.
Even with the convergence, however, the predicted cohort fertility for 2005--16 is 
above the 1992--93 fertility rate for every group except urban women with no education.
Furthermore, for the last period, the predicted cohort fertility is still between ten
and 20 percent higher than the fertility rate.

Another indication of the fertility rate bias is that it is increasing over time for some 
groups.
For example, urban women with eight to eleven years of education showed 1.84, 1.81, and
1.87 over the last three surveys.
These increases likely arise from the stabilization of the age of first birth and the 
spacing between births.

%
% \footnote{
% Increases in TFR over time, even in the face of declining cohort fertility, are not 
% unusual, as shown by Columbia in the early 1990s \citep{Bongaarts1999}.
% }

Despite the declines in the predicted cohort fertility, it is still mostly above the 
generally accepted replacement level of 2.1.
Only for urban women with twelve or more years of education is the predicted cohort
fertility clearly below 2.1, and even then, it is still more than 0.3 children higher than 
the fertility rate estimate of 1.5.
Finally, the predicted cohort fertility numbers are likely too low,
especially for the lower education groups, since it relies only on the
first four births and because the model ignores births that took place past the end
of the imposed 105 month birth interval censoring.

>>>>

% For word counting uncomment line below and raggedright below
% http://countwordsfree.com/#
\usepackage[nolists]{endfloat}

@@

% For word counting remove comment on the line below
\raggedright

@@

% \begin{figure}
% \centering
% \includegraphics[width=\textwidth,height=\textheight,keepaspectration=true]{bs_low_urban}
% \caption{combined graphs - low urban}
% \label{fig:spacing_low_urban}
% \end{figure}

@@

% \begin{figure}
% \centering
% \includegraphics[width=\textwidth,height=\textheight,keepaspectration=true]{bs_highest_rural}
% \caption{combined graphs - highest rural}
% \label{fig:spacing_highest_rural}
% \end{figure}

@@

The tempo effect from longer birth intervals means that the total fertility rate may 
underestimate cohort fertility. 
The next question I address is, therefore, to what extent did the changes bias the 
fertility estimates for India? 
To this end, I compare fertility based on a variation of the total fertility rate with 
predicted cohort fertility from the hazard model.
Table \ref{tab:fertility} shows the two fertility measures by urban/rural and education.

@@

The fertility rate follows the same procedure as in the Demographic and Health Survey 
reports: I use the births from 36 to 1 month before the survey month to calculate 
age-specific fertility rates for five-year age groups and then sum the age-specific 
fertility rates multiplied by five  \citep{Croft2018}.
However, because the hazard model predictions only use births up to parity four, I use the 
same set of births for the fertility rate and label it the ``four-parity'' fertility rate. 
Hence, the presented fertility rates are not directly comparable to those in the NFHS reports.

@@

Because NFHS-1 was after the introduction of sex selection, I cannot calculate a fertility 
rate in precisely the same manner for a period before sex selection was widely available.
Instead, I calculate the fertility rates for women between 15 and 39 years of age five years 
before the survey month, again using the number of births three years before.
This rate is shown as ``1987--88'' in the table.
Given the relatively low number of births to women 40 to 45 years of age, this approach 
provides the best estimate of the fertility rate when sex selection still was not 
widespread.

To predict cohort fertility based on the hazard models, I estimate the parity progression 
probability for each spell. 
Since parity progression depends on the sex composition of prior children, I estimate the 
probability for each sex composition and weigh the probabilities with the likelihood of 
the sex compositions.
The survey rounds do not coincide directly with the periods used for the hazard model.
Therefore, I compare the model results for 1972--1984, 1985--1994, 1995--2004,
and 2005--2016 with rounds 1 through 4 of the NFHS, respectively.

I include the spell from marriage to first birth, despite the problems capturing the exact 
timing of marriages since the estimated progression probabilities should not be affected 
by this problem. 
I begin with the age of marriage for each woman and predict the likelihood of progressing 
to each parity, assuming three years increases in age between births. 
Shorter assumed increases in age lead to slightly higher predicted fertility.

Sterilizations are not incorporated into the hazard model because most occur
immediately after giving birth. 
To compensate, I estimate the probability of sterilization using a Logit model and use 
that to scale down the parity progression probability when predicting cohort fertility.

The predicted cohort fertility based on the hazard model is higher than the four-parity 
fertility rate in almost all cases. 
Only women with no education in the first period show little difference between the two 
fertility measures, a situation where fertility is high, spacing very short, and likely 
unchanged for an extended period.

Consistent with a more substantial bias in the fertility rate when the age of marriage 
and the length of birth intervals increase, the absolute bias is least in the first 
and the last period and highest in the middle two periods.
Hence, the fertility rate declined too fast from the mid-1980s to the century's end.
Only recently, as the rate of increase for the birth intervals has slowed, do the two 
fertility measures begin to converge again.
Even with the convergence, the predicted 2005--16 cohort fertility is still above the 
1992--93 fertility rate for every group, except urban women with no education. 
Furthermore, for the last period, the predicted cohort fertility remains at least ten to 
twenty percent higher than the fertility rate.

Another indication of how tempo effects bias the fertility rate bias is that the
fertility rate \emph{increases} for some groups.
For example, for urban women with eight to eleven years of education, the fertility
rates were 1.84, 1.81, and 1.87 over the last three surveys.
This pattern likely arises from the stabilization of the age of first birth and the 
spacing between births.

Finally, even with the declines in the predicted cohort fertility, it is still mostly 
above replacement.
Only for urban women with twelve or more years of education is the predicted cohort
fertility clearly below 2.1.
Even then, cohort fertility is still more than 0.3 children higher than 
the fertility rate estimate of 1.5.
Furthermore, the predicted cohort fertility numbers are likely too low since I use only 
the first four births and births before the imposed 105-month birth interval censoring.
